Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-23 paper.tex

The classical approach to generating simulation data is
to render the results in some text format. For example,
\ms~\citep{hudson2002generating} outputs a set of sequences
and can also optionally output the marginal trees along
the genome in Newick format,
and variants of this approach are used by many simulators.
Text files have many advantages, but are slow to process
at scale. With sample sizes becoming ever larger, processing
text output can become a significant bottleneck in many analyses.
The ability to efficiently process simulation results is
particularly important in simulation-based inference methods
such as Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and machine learning based
approaches~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable,
sanchez2020deep}. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence
data, and methods avoid the bottleneck of parsing
text-based file formats to compute these statistics
by either developing their own
simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
or creating forked versions of existing
simulators~\cite[e.g.][]{thornton2006approximate,
hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
tightly integrated with the inference method.
While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error (and propagation of bugs).
Modern approaches to ABC such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} use large
numbers of weakly informative statistics,
making the need to efficiently compute statistics from simulation
results even more acute.

@@

% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
Even if we consider a single tree, \msprime\ generates mutations
extremely efficiently.
For example, we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim$260,000 mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods because it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms
at this scale. It seems quite unlikely, however,
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.

@@

The Sequentially Markovian Coalescent (SMC) is an approximation of the
coalescent with recombination
model~\citep{mcvean2005approximating,marjoram2006fast},
and was originally motivated to a significant degree by the need to
simulate longer genomes than was possible using tools like \ms.
The SMC is a good approximation to the
coalescent with recombination when we have four or fewer sampled
genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
effects of the approximation are less well understood for larger
sample sizes and several approaches have been proposed to
allow simulations more closely approximate the coalescent
with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
The SMC is an important analytical approximation
and has allowed the development of numerous inference
methods~\citep{
li2011inference,
harris2013inferring,
sheehan2013estimating,
schiffels2014inferring,
carmi2014renewal,
rasmussen2014genome,
zheng2014bayesian,
terhorst2017robust}.

@@

The SMC~\citep{mcvean2005approximating} and
SMC'~\citep{marjoram2006fast} models are available
in \msprime\ 1.0. However, they are currently implemented using a
naive rejection sampling approach, and are somewhat slower
to simulate than the exact coalescent with recombination. These
models are therefore only appropriate for studying the SMC approximations
themselves; an important avenue for future work is to implement
efficient SMC-like models, perhaps based on the approaches used
in \texttt{cosi2}~\citep{shlyakhter2014cosi2}, to facilitate
simulation of organisms such as \textit{Drosophila melagaster}.

@@

%introduction
Gene conversion is a unidirectional transfer of a short segment of genetic material,
for example between homologous chromosomes~\citep{chen2007gene}.
%description of ancestral process
Just as recombination, gene conversion events split the ancestral material
for some lineage, creating two independent lineages.
In contrast to the previously described recombination process, where the ancestral
material to the left and right of a recombination breakpoint follows two
different ancestral lineages, gene conversion events distribute the genetic material
of a short segment to one ancestral lineage and the part to the left and right of this
segment to another lineage.
%Franz Baumdicker: Not sure if this next sentence is necessary,
% but it might be of interest for some people?
The resulting distribution of ancestral material after a gene conversion event
equals the distribution after two recombination events at the left and right position
of the gene conversion tract, followed by a common ancestor event.
%Impact on LD
However, gene conversions impact much shorter segments than crossover recombinations,
typically below 1kb, and thus affects linkage pattern differently than
recombination~\citep{korunes2017gene}.

@@

Gene conversion is particularly useful to model homologous recombination in
bacterial evolution.
We compare the performance of \msprime\ with gene conversion to
SimBac~\citep{brown2016simbac}, a simulator developed specifically to
simulate bacterial evolution. Although \msprime\ is missing some features
relevant to this case, e.g.\ the simulation of circular genomes, it is far
more efficient at simulating the basic processes.
Furthermore, gene conversion can be considered in combination
with other current and future features of \msprime\ 1.0,
including varying recombination rates and complex demographies.

@@

\subsection*{Demography}
% TODO should we put in some citations here for the models? Hardly seems
% necessary.
One of the key applications of population genetic simulations is to generate
data for complex demographies. Beyond idealised cases such as stepping-stone or
island models, or specialised cases such as isolation-with-migration models,
analytical results are rarely possible. Simulation is therefore integral to the
development and evaluation of methods for demographic inference. The demography
model in \msprime\ is directly derived from the approach used in \ms, and
supports an arbitrary number of randomly mating populations exchanging migrants
at specified rates. A range of demographic events are provided, which allow for
varying population parameters and migration rates over time, as well as
population splits, admixtures and pulse migrations. The location of lineages
over time can be tracked in as much detail as required: each tree node is
automatically associated with the population in which it arose, the location of
lineages can be recorded at any given time via census events, or every lineage
migration can be recorded.

A major change for \msprime\ 1.0 is the introduction of the new Demography API,
designed to address a serious flaw in the \msprime\ 0.x interface which lead to
a number of avoidable errors in downstream
simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
separate parameters be provided to the \texttt{simulate} function to describe a
demographic model, making it easy to accidentally omit information. The 1.0 API
resolves this issue by creating a new \texttt{Demography} class, which
encapsulates all information about the demographic model, and fully decouples
the definition from other simulation details. An instance of this class is then
provided as a parameter to the new \texttt{sim\_ancestry} function,
substantially reducing the potential for error. Another improvement over the
0.X APIs is the introduction of explicit population split and admixture events,
and a population state machine that ensures that lineages cannot migrate into
(or be sampled from) inactive populations. This demography model is compatible
with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
class supports importing Demes models. We also support importing demographic
models in the form of Newick species tree descriptions, as well as those
produced by programs such as *BEAST~\citep{heled2009bayesian}.

Many popular approaches in population genetics use the distribution of
coalescence rates between pairs of lineages in one or more populations to infer
effective population sizes over
time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
split times and subsequent migration rates between
populations~\citep{wang2020tracking}. The \texttt{DemographyDebugger} provides
methods to compute a suite of numerical predictions about the location of
lineages over time, for example the coalescence rates for two or more lineages
drawn from populations at specified times in the past. This provides a valuable
ground-truth when evaluating such inference methods, as illustrated by
\cite{adrion2020community}.

@@

\begin{figure}
\includegraphics[width=6cm]{figures/sweeps-perf}
\caption{\label{fig-selection-perf}Comparison of the runtime for
\msprime\ and \texttt{discoal} for simulating selective sweeps.
}
\end{figure}

>>>>

The classical approach to generating simulation data is
to render the results in some text format. For example,
\ms~\citep{hudson2002generating} outputs a set of sequences
and can also optionally output the marginal trees along
the genome in Newick format,
and variants of this approach are used by many simulators.
Text files have many advantages, but are slow to process
at scale. With sample sizes becoming ever larger, processing
text output can become a significant bottleneck in many analyses.
The ability to efficiently process simulation results is
particularly important in simulation-based inference methods
such as Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and machine learning based
approaches~\citep{sheehan2016deep,chan2018likelihood,schrider2018supervised,
flagel2019unreasonable,sanchez2020deep}. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which one can sample from the model and parameter space.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence
data, and methods avoid the bottleneck of parsing
text-based file formats to compute these statistics
by either developing their own
simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
or creating forked versions of existing
simulators~\cite[e.g.][]{thornton2006approximate,
hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
tightly integrated with the inference method.
While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error (and propagation of bugs).
Modern approaches to ABC such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} use large
numbers of weakly informative statistics,
making the need to efficiently compute statistics from simulation
results even more acute.

@@

% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
Even if we consider a single tree, \msprime\ generates mutations
extremely efficiently.
For example, we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim$260,000 mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods because at this scale it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms.
It seems quite unlikely, however,
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.

@@

The Sequentially Markovian Coalescent (SMC) is an approximation of the
coalescent with recombination
model~\citep{mcvean2005approximating,marjoram2006fast},
and was originally motivated to a significant degree by the need to
simulate longer genomes than was possible using tools like \ms.
The SMC is a good approximation to the
coalescent with recombination when we have four or fewer sampled
genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
effects of the approximation are less well understood for larger
sample sizes and several approaches have been proposed to
allow simulations that more closely approximate the coalescent
with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
The SMC is an important analytical approximation
and has allowed the development of numerous inference
methods~\citep{
li2011inference,
harris2013inferring,
sheehan2013estimating,
schiffels2014inferring,
carmi2014renewal,
rasmussen2014genome,
zheng2014bayesian,
terhorst2017robust}.

@@

The SMC~\citep{mcvean2005approximating} and
SMC'~\citep{marjoram2006fast} models are available
in \msprime\ 1.0. However, they are currently implemented using a
naive rejection sampling approach, and are somewhat slower
to simulate than the exact coalescent with recombination. These
models are therefore only appropriate for studying the SMC approximations
themselves; an important avenue for future work is to implement
efficient SMC-like models, perhaps based on the approaches used
in \texttt{cosi2}~\citep{shlyakhter2014cosi2}, to facilitate
simulation of organisms such as \textit{Drosophila melanogaster}.

@@

%introduction
Gene conversion is a unidirectional transfer of a short segment of genetic material,
for example between homologous chromosomes~\citep{chen2007gene}.
%description of ancestral process
Just as recombination, gene conversion events split the ancestral material
for some lineage, creating two independent lineages.
In contrast to the previously described recombination process, where the ancestral
material to the left and right of a recombination breakpoint follows two
different ancestral lineages, gene conversion events distribute the genetic material
of a short segment to one ancestral lineage and the part to the left and right of this
segment to another lineage.
%Franz Baumdicker: Not sure if this next sentence is necessary,
% but it might be of interest for some people?
The resulting distribution of ancestral material after a gene conversion event
equals the distribution after two recombination events at the left and right position
of the gene conversion tract, followed by a common ancestor event.
%Impact on LD
However, gene conversions impact much shorter segments than crossover recombinations,
typically below 1kb, and thus affect linkage patterns differently than
recombinations~\citep{korunes2017gene}.

@@

Gene conversion is particularly useful to model homologous recombination in
bacterial evolution,
so we compare the performance of \msprime\ with gene conversion to
SimBac~\citep{brown2016simbac}, a simulator developed specifically to
simulate bacterial evolution. Although \msprime\ is missing some features
relevant to this case, e.g.\ the simulation of circular genomes, it is far
more efficient at simulating the basic processes.
Furthermore, gene conversion can be considered in combination
with other current and future features of \msprime\ 1.0,
including varying recombination rates and complex demographies.

@@

\subsection*{Demography}
% TODO should we put in some citations here for the models? Hardly seems
% necessary.
One of the key applications of population genetic simulations is to generate
data for complex demographies. Beyond idealised cases such as stepping-stone or
island models, or specialised cases such as isolation-with-migration models,
analytical results are rarely possible. Simulation is therefore integral to the
development and evaluation of methods for demographic inference. The demography
model in \msprime\ is directly derived from the approach used in \ms, and
supports an arbitrary number of randomly mating populations exchanging migrants
at specified rates. A range of demographic events are supported, which allow for
varying population sizes and growth rates, changing migration rates over time,
as well as population splits, admixtures and pulse migrations.
The location of sampled lineages
can be tracked through time in as much detail as required: each tree node is
automatically associated with the population in which it arose, the location of
lineages can be recorded at any given time via census events, or every lineage
migration can be recorded.

A major change for \msprime\ 1.0 is the introduction of the new Demography API,
designed to address a serious flaw in the \msprime\ 0.x interface which lead to
a number of avoidable errors in downstream
simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
separate parameters be provided to the \texttt{simulate} function to describe a
demographic model, making it easy to accidentally omit information. The 1.0 API
resolves this issue by creating a new \texttt{Demography} class, which
encapsulates all information about the demographic model, and fully decouples
the definition from other simulation details. An instance of this class is then
provided as a parameter to the new \texttt{sim\_ancestry} function,
substantially reducing the potential for error.
Another improvement over the
0.X APIs is the introduction of explicit population split and admixture events,
and a population state machine that ensures that lineages cannot migrate into
(or be sampled from) inactive populations. This demography model is compatible
with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
class supports importing and exporting Demes models.
Models previously constructed using the 0.x API can be seamlessly imported into
the \texttt{Demography} class, and we also support importing demographic
models in the form of Newick species tree descriptions, as well as those
produced by programs such as *BEAST~\citep{heled2009bayesian}.

The \texttt{DemographyDebugger} can output detailed information for debugging
demographic models, but also provides
methods to compute a suite of numerical predictions about the location of
lineages over time, for example the coalescence rates for two or more lineages
drawn from populations at specified times in the past.
Many popular approaches in population genetics use the distribution of
coalescence rates between pairs of lineages in one or more populations to infer
effective population sizes over
time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
split times and subsequent migration rates between
populations~\citep{wang2020tracking}.
The \texttt{DemographyDebugger} provides a valuable
ground-truth when evaluating such inference methods, as illustrated by
\cite{adrion2020community}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/sweeps-perf}
\end{center}
\caption{\label{fig-selection-perf}Comparison of the runtime for
\msprime\ and \texttt{discoal} for simulating selective sweeps.
}
\end{figure}
